\vspace{-2.0em} Move Schedules: \\ Fast persistence computations in sparse dynamic settings\thanks{This work was partially supported by the National Science Foundation through grants CCF-2006661 and CAREER award DMS-1943758.}\vspace{-0.5em}

### 2021-08-04 move_schedules.tex
Reworking the last section on image patches results

 \subsubsection*{Natural images dataset} 
 In the previous section, we outlined the computational theory of multidimensional persistence, including its relevance to our proposed move scheduling approach. In this section, we demonstrate the efficiency of our method on practical use case of multidimensional persistence: detecting the presence of a low-dimensional topological space which well-approximates the distribution of high dimensional data.

 \begin{figure}[t]
	\includegraphics[width=0.98\textwidth]{natural_images}
	\caption{Bipersistence example of natural images data set on an $12 \times 16$ coarsened grid. On the left, a projection the full data set onto the first two DCT basis vectors is shown, along with the 15 landmark patches. In the middle, the bigraded Betti numbers, the Hilbert function, and a chosen line $L$ are shown for the coarsened grid. As before, the $0/1/2$ dimension bigraded Betti numbers are shown in green/red/yellow, respectively. The blue regions highlights where $\mathrm{dim}(M) = 5$. On the right, the persistence diagram of $L$, with larger points indicating a multiplicity of 2. Observe five persistent features are revealed in both the middle and right plots, matching $\beta_1$ of the three-circle model. }
	\label{fig:patch_data_dgm}
\end{figure}

A common hypothesis is that high dimensional data tend to lie in the vicinity of an embedded, low dimensional manifold or topological space. An exemplary demonstration of this is given in~\cite{lee2003nonlinear}, who explored the space of $3 \times 3$ high-contrast patches extracted from Van Hatern's still image collection, which consists of approximately 4,000 monochrome images depicting various areas outside around Groningen (Holland)~\cite{}. The study of how empirical data from natural images is distributed with respect to predicted spaces and manifolds is motivated by applications in sparse coding and efficient image representations. In particular, Lee et al.~\cite{} was interested in analyzing the empirical probability distribution of high-contrast $3 \times 3$ patches of natural images, in pixel-space, towards understanding whether there were any clear qualitative differences the distributions of patches extracted from from images of different modalities, e.g. optical versus range images. 
Experiments conducted by Carllsson et al.~\cite{} using persistent homology confirmed that the distribution of high-contrast $3 \times 3$ patches is indeed well-approximated by a Klein bottle---around 60\% of the high-contrast patches from the still image data set are concentrated within a small neighborhood around a model of the Klein bottle which accounts for $\approx$ 21\% of the volume $S^7$, compared to the $\approx$ 84\% of volume of $S^7$ accounted for by the entire data set. In a subsequent effort by Perea et al.~\cite{}, a framework for estimating and representing the distribution around low dimensional submanifolds of pixel space is given, establishing a principled way of obtaining multi-scale rotation invariant descriptors useful for classification tasks.

Using the same high-contrast patch data set studied in~\cite{lee2003nonlinear}, we evaluate the performance of various methods at computing the rank invariant using the parameterization from~\ref{sec:fibered_barcode_reparam}. Specifically, we begin by extracting a representative subset of the set $X \subset \mathcal{X}$ of patches representing the top 30\% densest patches of the space of all patches $\mathcal{X}$. Carlsson et al.~\cite{} studied a variety of choices of $k$ for a k-nearest neighbor (knn) density estimator $\rho$---which determines which points are the ``densest''---in order to understand how the data were distributed around the Klein bottle model. It was determined empirically that when $k \approx 15$, the top 30\% captures three types of directional shifts in the grayscale gradients of the image patches. These three gradients were shown to be captured well by the three-circle model $C_3$---the 1-skeleton of the Klein bottle. In this space, there are three circles, two of which (say, $S_v$ and $S_h$) intersect the third (say, $S_\mathrm{lin}$) in exactly two points, but they themselves do not intersect. One of the circles, $S_{\mathrm{lin}}$, parameterizes a the rotational gradient, whereas the other two $S_v$ and $S_h$ parameterize vertical and horizontal gradients. A projection of the data $X$ onto the first two DCT basis, is shown in the left-most image in Figure~\ref{fig:patch_data_dgm}, along with 15 landmark points. It was shown in~\cite{} that when $k \approx 300$, the top 30\% of data extracted from the knn density estimator $\rho_{300}$ `fills out' the two-manifold, yielding Betti numbers $\beta_0 = \beta_1 = 1$, whereas when $k$ is decreases to a more localized estimate, the three-circle model is recovered, which has Betti numbers $\beta_0 = 1, \beta_1 = 5$.

We sampled landmarks from $X \subset \mathcal{X}$ using a stratified sampling approach with $n = 25$ strata, where each strata contains the $(1/n)$-thick level set given by the knn density estimate $\rho_{15}$. The use of landmarks gives us certain coverage guarantees that the geometry is preserved within each level set, whereas the stratification ensures the original density of is approximated preserved as well. From this data set, we construct a Rips-density bifiltration $\mathcal{F}$ using $\rho_{15}$ equipped with the geodesic metric computed over the $k = 15$ knn graph of $X$. In theory, this bifiltration should be well approximated by $C_3$, which we ought to be able to determine by a careful inspection of the rank invariant. That is, since $C_3$ has $\beta_1  = 5$, there should be some line $L$ which induces a 1-parameter filtration whose corresponding persistence diagram (for dimension $1$) displays five highly persistent features. We show an example of the data, the bigraded Betti numbers, and the persistence diagram extracted from a particular choice of line $L$ in Figure~\ref{fig:patch_data_dgm}.

@@

\begin{table}[h]
\caption{Cost to compute all barcode templates $\mathcal{T}$ }\label{tab:barcode_templates}
\centering
\begin{tabular}{ |P{1.6cm}|P{1.6cm}|P{1.2cm}|P{2.8cm}|P{2.4cm}|}
 \hline
 %\multicolumn{7}{|c|}{ Size and cost of constructing $\mathcal{A}^\ast(M)$ } \\
 %\hline
  \multicolumn{2}{|c|}{ $\beta$ and $\mathcal{A}(M)$ } & \multicolumn{3}{c|}{ Column reductions / Permutations } \\
 \hline
 Coarsening & \# 2-cells & \texttt{phCol} & Vineyards / $d_{K}$ & Moves / $d_{\text{LCS}}$ \\
 \hline
 8 x 8 & 39 & 94.9K & 245K / 1.53M & 38.0K / 11.6K   \\
 \hline
 12 x 12 & 127 & 318K & 439K / 2.66M & 81.9K / 33.0K \\
 \hline 
 16 x 16 & 425 & 1.07M & 825K / 4.75M & 114K / 87.4K \\
 \hline 
  20 x 20 & 926 & 2.32M & 1.15M / 6.77M  & 148K / 154K\\
 \hline 
  24 x 24 & 1.53K & 3.92M & 1.50M / 8.70M & 184K / 232K  \\
   \hline
 \end{tabular}
\label{table:dual_graph_costs}
\end{table}

>>>>

 \subsubsection*{Natural images dataset} 
 In the previous section, we outlined the computational theory of multidimensional persistence, including its relevance to our proposed move scheduling approach. In this section, we demonstrate the efficiency of our method on practical use case of multidimensional persistence: detecting the presence of a low-dimensional topological space which well-approximates the distribution of texture patches extracted from real-world image data.

 \begin{figure}[t]
	\includegraphics[width=0.98\textwidth]{natural_images}
	\caption{Bipersistence example of natural images data set on an $12 \times 16$ coarsened grid. On the left, a projection the full data set is shown, along with the 15 landmark patches. In the middle, the bigraded Betti numbers, the Hilbert function, and a chosen line $L$ are shown for the coarsened grid. As before, the $0/1/2$ dimension bigraded Betti numbers are shown in green/red/yellow, respectively. The blue regions highlights where $\mathrm{dim}(M) = 5$. On the right, the persistence diagram of $L$, with larger points indicating a multiplicity of 2. Observe five persistent features are revealed in both the middle and right plots, matching $\beta_1$ of the three-circle model. }
	\label{fig:patch_data_dgm}
\end{figure}

A common hypothesis is that high dimensional data tend to lie in the vicinity of an embedded, low dimensional manifold or topological space. An exemplary demonstration of this is given in~\cite{lee2003nonlinear}, who explored the space of $3 \times 3$ high-contrast patches extracted from Hans van Hateren's still image collection~\cite{hateren_schaaf_1998}, which consists of approximately 4,000 monochrome images\footnote{See \url{http://bethgelab.org/datasets/vanhateren/} for details on the image collection.} depicting various areas outside around Groningen (Holland). The study of how empirical data extracted from natural images is distributed with respect to predicted spaces and manifolds is motivated by applications in sparse coding and efficient image representations. In particular, Lee et al.~\cite{lee2003nonlinear} was interested in analyzing the empirical probability distribution of high-contrast $3 \times 3$ patches of natural images, in pixel-space, towards understanding whether there were any clear qualitative differences the distributions of patches extracted from from images of different modalities, e.g. optical versus range images. 
Experiments conducted by Carllsson et al.~\cite{carlsson2008local} using persistent homology confirmed that the distribution of high-contrast $3 \times 3$ patches is indeed well-approximated by a Klein bottle---around 60\% of the high-contrast patches from the still image data set are concentrated within a small neighborhood around a model of a Klein bottle which accounts for $\approx$ 21\% of the volume $S^7$, compared to the $\approx$ 84\% of volume of $S^7$ accounted for by the entire data set. Along a similar vein but from a sparse coding perspective, Perea et al.~\cite{perea2014klein} introduced dictionary learning framework for estimating and representing the distribution of patches from texture images.

A priori, had one not been aware of the analysis done by~\cite{lee2003nonlinear, hateren_schaaf_1998, carlsson2008local, perea2014klein}, there is no specific reason to assume the data are well-approximated by the Klein bottle specifically.

Using the same high-contrast patch data set studied in~\cite{lee2003nonlinear}, we evaluate the performance of various methods at computing the fibered barcode invariant using the parameterization from~\ref{sec:fibered_barcode_reparam}. First, we give a summary of the data set. 
Specifically, we begin by extracting a representative subset of the set $X \subset \mathcal{X}$ of patches representing the top 30\% densest patches of the space of all patches $\mathcal{X}$, where the notion of density is approximated with a k-nearest neighbor (knn) density estimator $\rho$. Carlsson et al.~\cite{carlsson2008local} studied a variety of choices of $k$ in order to understand how localized the data were distributed around the Klein bottle model. He determined empirically that when $k \approx 15$, the top 30\% captures three types of directional shifts in the grayscale gradients of the image patches. These three gradients were shown to be captured well by the three-circle model $C_3$---the 1-skeleton of the Klein bottle. The $C_3$ model consists of three circles, two of which (say, $S_v$ and $S_h$) intersect the third (say, $S_\mathrm{lin}$) in exactly two points, but themselves do not intersect. One of the circles, $S_{\mathrm{lin}}$, parameterizes a the rotational gradient, whereas the other two $S_v$ and $S_h$ parameterize vertical and horizontal gradients.  
A projection of the data $X$ is shown in the left-most image in Figure~\ref{fig:patch_data_dgm}, along with 15 landmark points. It was shown in~\cite{carlsson2008local} that when $k \approx 300$, the top 30\% of data extracted from the knn density estimator $\rho_{300}$ `fills out' the two-manifold, yielding Betti numbers $\beta_0 = \beta_1 = 1$, whereas when $k$ is decreases to a more localized estimate, the three-circle model is recovered, which has Betti numbers $\beta_0 = 1, \beta_1 = 5$.

Due to the aformentioned high complexity of the fibered barcode computation, we begin by working with a subset of the image patch data $\mathcal{X}$. Specifically, we combine the use furthest-point sampling and proportionate allocation (stratified) sampling to sample landmarks $X \subset \mathcal{X}$ distributed within $n = 25$ strata. Each strata consists of the $(1/n)$-thick level set given by $\rho_{15}$, the density estimator. The use of furthest-point sampling gives us certain coverage guarantees that the geometry is approximately preserved within each level set, whereas the stratification ensures the original density of is approximated preserved as well. From this data set, we construct a Rips-(co)density bifiltration $\mathcal{F}$ using $\rho_{15}$ equipped with the geodesic metric computed over the $k = 15$ knn graph on $X$.

In theory, this bifiltration should be well approximated by the $C_3$ model, which one ought to be able to determine by a careful inspection of the fibered barcode. That is, since $C_3$ has $\beta_1  = 5$, there should be some line $L$ which induces a 1-parameter filtration whose corresponding persistence diagram (for dimension $1$) displays five highly persistent features. We show an example of the data, the bigraded Betti numbers, and the persistence diagram extracted from a particular choice of line $L$ in Figure~\ref{fig:patch_data_dgm}.

@@

\begin{table}[h]
\caption{Cost to compute all barcode templates $\mathcal{T}$ for various coarsening choices $\beta(M)$.}\label{tab:barcode_templates}
\centering
\begin{tabular}{ |P{1.6cm}|P{1.6cm}|P{1.2cm}|P{2.8cm}|P{2.4cm}|}
 \hline
 %\multicolumn{7}{|c|}{ Size and cost of constructing $\mathcal{A}^\ast(M)$ } \\
 %\hline
  \multicolumn{1}{|c|}{ $\beta(M)$ } & \multicolumn{1}{|c|}{$\mathcal{A}(M)$ } & \multicolumn{3}{c|}{ Column reductions / Permutations } \\
 \hline
 Coarsening & \# 2-cells & \texttt{phCol} & Vineyards / $d_{K}$ & Moves / $d_{\text{LCS}}$ \\
 \hline
 8 x 8 & 39 & 94.9K & 245K / 1.53M & 38.0K / 11.6K   \\
 \hline
 12 x 12 & 127 & 318K & 439K / 2.66M & 81.9K / 33.0K \\
 \hline 
 16 x 16 & 425 & 1.07M & 825K / 4.75M & 114K / 87.4K \\
 \hline 
  20 x 20 & 926 & 2.32M & 1.15M / 6.77M  & 148K / 154K\\
 \hline 
  24 x 24 & 1.53K & 3.92M & 1.50M / 8.70M & 184K / 232K  \\
   \hline
 \end{tabular}
\label{table:dual_graph_costs}
\end{table}
